---
title: "Class 11 Pt. 1: RNASeq Galaxy"
author: "Katherine Lim (A15900881)"
format: pdf
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=105535077;sample=MXL#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
round(table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100, 2)
```

Now let's look at a different population (GBR).

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Let's dig into this further.

# Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale.

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(dplyr)

median_expression <- expr %>%
  group_by(geno) %>%
  summarize(median_expression = median(exp))

sample_size <- expr %>%
  count(geno)

combined_data <- merge(median_expression, sample_size, by = "geno")
combined_data

```

How many samples do we have?

```{r}
nrow(expr)
```

```{r}
library(ggplot2)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? 

Let's make a boxplot

```{r}
ggplot(expr) + 
  aes(x = geno, y = exp, fill = geno) + 
  geom_boxplot(notch = TRUE)
```

A/A is expressed more than G/G. 










