---
title: "Class 11: Halloween Candy Mini-Project"
author: "Katherine Lim (A15900881)"
format: pdf
---

In today's class we will examine candy data and see if this helps us gain some more feeling for how PCA and other methods work.

```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

> Q. What are these fruity candies?

```{r}
rownames(candy[candy$fruity == 1, ])
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Strawberry bon bons", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

"Winpercent" is on a 0:100 scale while the rest of the variables are on a 0:1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero represents that the candy does not have chocolate in it while a one represents that the candy has chocolate in it.

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins = 10)
```

> Q9. Is the distribution of winpercent values symmetrical?

No.

> Q10. Is the center of the distribution above or below 50%?

It is below 50%.

```{r}
mean(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

To answer this questions I will need to:
- "subset" (a.k.a select or filter) the candy dataset to just chocolate candy 
- get their winpercent values
- and then calculate the mean of these
Then do the same for fruity candy and compare.

```{r}
chocolate.winpercent <- mean(candy[candy$chocolate == 1, ]$winpercent)
chocolate.winpercent

fruity.winpercent <- mean(candy[candy$fruity == 1, ]$winpercent)
fruity.winpercent
```

Chocolate candy is higher ranked than fruity candy. 

> Q12. Is this difference statistically significant?

```{r}
t.test((candy[as.logical(candy$chocolate), "winpercent"]), (candy[as.logical(candy$fruity), "winpercent"]))
```
## Overall candy rankings

There is a base R function called `sort()` for sorting vectors of input.

```{r}
x <- c(2, 5, 10)

sort(x)
```

The `order()` function can be used to return the indices of the inpurt that would result in it being sorted. 

> Q13. What are the five least liked candy types in this set?

```{r}
disliked_candy <- candy[order(candy$winpercent, decreasing = FALSE), ]
head(disliked_candy, n = 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
top_candy <- candy[order(candy$winpercent, decreasing = TRUE), ]
head(top_candy, n = 5)
```

> Q15 & Q16. Make a first barplot of candy ranking based on winpercent values.

```{r}
my_cols = rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill = my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets.

> Q18. What is the best ranked fruity candy?

Starburst.

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label = rownames(candy)) +
  geom_point(col = my_cols) + 
  geom_text_repel(col = my_cols, size = 3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

```{r}
ord <- order(candy$winpercent, decreasing = TRUE)
head(candy[ord, c(11, 12)], n = 5)
```
Reese's Miniatures.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head(candy[ord, c(11, 12)], n = 5)
```

Nik L Nip.

## Exploring the correlation structure

Pearson correlation goes between -1 and +1 with zero indicating no correlation. Values close to one indicate high correlation.

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar or winpercent.

### Principal Component Analysis

The base R function for PCA is called `prcomp()` and we can set `scale() = TRUE/FALSE`.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

The main result of PCA - i.e. the new PC plot (projection of candy on our new PC axis) is contained in `pca$x`.

```{r}
pc <- as.data.frame(pca$x)

ggplot(pc) +
  aes(PC1, PC2, label = rownames(pc)) +
  geom_point(col = my_cols) +
  geom_text_repel(max.overlaps = 5)
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard and pluribus.
